# Week 5


This week we have used the data from United Nations Development Programme, Human Development Reports ([hdr.undp.org](http://hdr.undp.org/en/content/human-development-index-hdi)). 

From the reports, the two datasets of Human Development Index (HDI) and Gender Inequality Index were downloaded, combined, and some data variables were selected for this week's analysis exercise.

```{r}
# Load the human.txt data into R.

human<-read.table("~/Desktop/IODS-project/data/human.txt", sep="", header=T)

# Explore the structure and the dimensions of the data 
str(human)
dim(human)
```

In our dataset we have 155 objects and 8 variables, so we have 155 rows (different countries) and 8 columns. All the variables are recogniced as numerical, most of them have decimal values, but the GNI and Mat.Mor are integers. The chosen 8 variables (column heading abbreviations) from larger data set describe following things:


1. The proportion of females divided with proportion of males with at least secondary education (Edu2.FM).
2. The proportion of females divided with the proportion of males in the labour force. (Labo.FM)
3. Expected years of schooling for children of school entering age (Edu.Exp)
4. Life expectancy at birth (Life.Exp)
5. Gross national income per capita (GNI) 
6. Maternal mortality ratio (Mat.Mor)
7. Adolescent birth rate (Ado.Birth)
8. Percetange of female representatives in parliament (Parli.F)

```{r}
# Show a graphical overview of the data
# Access to ggplot2 and tidyr
library(ggplot2)
library(tidyr)

# For easier overview of the data, the data set is scaled and class changed from numeric to data frame.
human_scaled<-scale(human)
human_scaled<-as.data.frame(human_scaled)

# Next scaled data is gathered to key-value pairs and plotted for visualization of the data.
gather(human_scaled)%>%ggplot(aes(value))+facet_wrap("key", scales="free_y", )+geom_histogram(bins=155, col="darkblue")+xlim(-6,6)

# To see how scaling affected another histograms are made for the original data.
gather(human)%>%ggplot(aes(value))+facet_wrap("key", scales="free", )+geom_histogram(bins=155, col="darkred")


```

From the blue histograms (scaled values) it is perhaps easier to see, which variables are normally distributed, and which are not. Ado.Birth, GNI and Mat.Mor are not following the normal distribution. From red histograms (not-scaled histograms) it is obvious that dataset doesn't have negative values, which makes sense, because none of the variables in the data table can have negative value. 

The summary tables below are first for not-scaled data and the next one is for the scaled data. These two tables (and histograms above) show that largest distribution is with GNI variable and least distributed is Life.Exp.

```{r}
# Creating summary tables for both scaled and original human data
summary(human)
summary(human_scaled)
```

The relationships between variables can be visualized by creating correlation plot. In this first correlation plot below, the distribution of the data variables is again visible in the middle of the plot (from higher left to lower bottom). The highest correlation value can be seen between variables Life.Exp and Edu.Exp. This is positively correlated, so in countries where expected years of schooling is high, the life expectancy is high also. Negatively correlated is the Mat.Mor and Life.Exp, so in those places, where life expectancy is high, the maternal mortality ratio is low. Least correlated were Edu2.FM and Labo.FM, so the proportion of females divided with proportion of males with at least secondary education and the proportion of females divided with the proportion of males in the labour force were not correlated. The same observations can be seen from the second correlation plot below.

```{r}
# Access GGally and corrplot
library(GGally)
library(corrplot)
# Visualize data variables with ggpairs()
ggpairs(human)

# Compute correlation matrix and visualize it with corrplot()
cor(human)%>%corrplot(order ="hclust", method="color", type="lower")

```

The not standardized human data is then analyzed with PCA (principal component analysis). The variability of this data is almost completely from component 1 (99.99% of variability) and the second component (orthologonal to PC1) makes the 0.01% variation in the data. The other principal components don't have importance in terms of captured variance. So, we have the two principal components (uncorrelated variables) which will capture the highest variation. The angles of arrows, however, are all looking the same, so this is not giving us much information about the correlation, because all the arrows are small angle with PC1 axis, meaning that the correlation is high between the features and PC1. The length of the arrow represents the standard deviation of the feature. 

```{r}
# Perform principal component analysis on the not standardized human data (with the SVD method)
pca_human<-prcomp(human)

# Create and print summary of pca_human
s<-summary(pca_human)
s

# Rounded percentages of variance captured by each PC, print them out
pca_pr<-round(1*s$importance[2,]*100, digits=2)
pca_pr

# Create object pc_lab to be used as axis labels
pc_lab<-paste0(names(pca_pr), "(", pca_pr, "%)")

# Draw a biplot (and suppress warnings for zero-lenght arrows)
suppressWarnings(biplot(pca_human, cex=c(0.8,1), col=c("grey40", "deeppink2"), xlab=pc_lab, ylab=NA, main="Biplot using not standardized human data"))

```

Next the PCA is done for the standardized variables. Now we can get more principal components. Comparing not-standardized and standardized PCAs, it is clear that PCA is sensitive to the scaling of original values, assuming that features that have larger variance are more important. From this standardized biplot it is clearly visible that Labo.FM and Parli.F are their own group, as well Mat.Mor and Ado.Birth are, and all the rest 4 variables are in the grouped on the left side of this biplot. So this means that maternal mortality ratio and 
adolescent birth rate are somehow related onto similar features and they are quite opposite of the high life expectancy at birth, expected years of schooling for children of school entering age, gross national income per capita and the proportion of females divided with proportion of males with at least secondary education. In other words, if we think about countries, those countries are more left which have higher education (also for girls), population health, and their people have more money to use. On the right are the countries which have higher mortality ratio for mother in childbirth and under aged girls are becoming mothers. On the top there are countries with larger number proportion of females in work life or having place in the parlament, while below part of the biplot there are less females than males in the parlament and work.


```{r}
# Perform principal component analysis on the not standardized human data (with the SVD method)
pca_human_scaled<-prcomp(human_scaled)

# Create and print summary of pca_human
s_s<-summary(pca_human_scaled)
s_s

# Rounded percentages of variance captured by each PC, print them out
pca_pr2<-round(1*s_s$importance[2,]*100, digits=1)
pca_pr2

# Create object pc_lab to be used as axis labels
pc_lab<-paste0(names(pca_pr2), "(", pca_pr2, "%)")

# Draw a biplot (and suppress warnings for zero-lenght arrows)
suppressWarnings(biplot(pca_human_scaled, cex=c(0.8,1), col=c("grey40", "deeppink2"), xlab=pc_lab, ylab=NA, main="Biplot using standardized human data"))
```
