# Week 5


This week we have used the data from United Nations Development Programme, Human Development Reports ([hdr.undp.org](http://hdr.undp.org/en/content/human-development-index-hdi)). 

From the reports, the two datasets of Human Development Index (HDI) and Gender Inequality Index were downloaded, combined, and some data variables were selected for this week's analysis exercise.

```{r}
# Load the human.txt data into R.

human<-read.table("~/Desktop/IODS-project/data/human.txt", sep="", header=T)

# Explore the structure and the dimensions of the data 
str(human)
dim(human)
```

In our dataset we have 155 objects and 8 variables, so we have 155 rows (different countries) and 8 columns. All the variables are recogniced as numerical, most of them have decimal values, but the GNI and Mat.Mor are integers. The chosen 8 variables (column heading abbreviations) from larger data set describe following things:


1. The proportion of females divided with proportion of males with at least secondary education (Edu2.FM).
2. The proportion of females divided with the proportion of males in the labour force. (Labo.FM)
3. Expected years of schooling for children of school entering age (Edu.Exp)
4. Life expectancy at birth (Life.Exp)
5. Gross national income per capita (GNI) 
6. Maternal mortality ratio (Mat.Mor)
7. Adolescent birth rate (Ado.Birth)
8. Percetange of female representatives in parliament (Parli.F)

```{r}
# Show a graphical overview of the data
# Access to ggplot2 and tidyr
library(ggplot2)
library(tidyr)

# For easier overview of the data, the data set is scaled and class changed from numeric to data frame.
human_scaled<-scale(human)
human_scaled<-as.data.frame(human_scaled)

# Next scaled data is gathered to key-value pairs and plotted for visualization of the data.
gather(human_scaled)%>%ggplot(aes(value))+facet_wrap("key", scales="free_y", )+geom_histogram(bins=155, col="darkblue")+xlim(-6,6)

# To see how scaling affected another histograms are made for the original data.
gather(human)%>%ggplot(aes(value))+facet_wrap("key", scales="free", )+geom_histogram(bins=155, col="darkred")


```

From the blue histograms (scaled values) it is perhaps easier to see, which variables are normally distributed, and which are not. Ado.Birth, GNI and Mat.Mor are not following the normal distribution. From red histograms (not-scaled histograms) it is obvious that dataset doesn't have negative values, which makes sense, because none of the variables in the data table can have negative value. 

The summary tables below are first for not-scaled data and the next one is for the scaled data. These two tables (and histograms above) show that largest distribution is with GNI variable and least distributed is Life.Exp.

```{r}
# Creating summary tables for both scaled and original human data
summary(human)
summary(human_scaled)
```

The relationships between variables can be visualized by creating correlation plot. In this first correlation plot below, the distribution of the data variables is again visible in the middle of the plot (from higher left to lower bottom). The highest correlation value can be seen between variables Life.Exp and Edu.Exp. This is positively correlated, so in countries where expected years of schooling is high, the life expectancy is high also. Negatively correlated is the Mat.Mor and Life.Exp, so in those places, where life expectancy is high, the maternal mortality ratio is low. Least correlated were Edu2.FM and Labo.FM, so the proportion of females divided with proportion of males with at least secondary education and the proportion of females divided with the proportion of males in the labour force were not correlated. The same observations can be seen from the second correlation plot below.

```{r}
# Access GGally and corrplot
library(GGally)
library(corrplot)
# Visualize data variables with ggpairs()
ggpairs(human)

# Compute correlation matrix and visualize it with corrplot()
cor(human)%>%corrplot(order ="hclust", method="color", type="lower")

```

